Bird beak's elaboration and exploration

### 2022-10-17 Analysis/11-variance_partitioning.Rmd
Update 11-variance_partitioning.Rmd

```{r}
library(dispRity)
library(mcmcmcglmmm)
library(dads)
```

# Variance explained in a simple model

Here I'm just doing a one dimensional model of a one dimensional trait (BM) and two clades (`"s"` and `"t"`).

We will now run it as the following model:

```
data ~ residuals + random.phylo + random.clade_s + random.clade_t
```

```{r, echo = FALSE, message = FALSE}
## Setting up the model
first_model <- make.mini.chains(data, dimensions = c(1:2), tree = tree, trait.family = "gaussian", residuals = "global", randoms = c("global", "clade"), verbose = TRUE, parameters =  list(nitt = 20000))

And this is what it looks like graphically

```{r, echo = FALSE}
## Model object
first_model_results <- MCMCglmm.subsets(
      data          = data,
      posteriors    = first_model_run,
      group         = MCMCglmm.levels(first_model_run)[c(1:3)],
      rename.groups = c("clade_s", "clade_t", "phylogeny"))

## Plotting the model
covar.plot(first_model_results, col = clade_cols, ellipses = mean, legend = TRUE)
```
We can then extract the variance of the model:

```{r, echo = FALSE}
#@ name plot.decompose.var
#@ description plotting the variance of a model in different ways
#@ param data a dispRity object
#@ param posteriors a MCMCglmm object
#@ param group.cols the colours for each term
#@ param obs.col the colour for the observed variance
#@ param relative whether to make the variance relative or absolute
#@ param plot which panels to plot. Can be any of the followings (default is all):
# "var boxplot" = variance per group;
# "var by n" = variance per sample size;
# "log var boxplot" = logged variance per group;
# "log var by n" = logged variance per sample size;
# "corrected var boxplot" = logged variance divided by logged sample sizel;
# "corrected var by n" = logged variance per logged sample size divided by logged sample size;
# "standard error" = variance per group divided by sqrt(n);
# "log standard error" = logged standard error;

plot.decompose.var <- function(data, posteriors, group.cols, obs.col = "red", relative = FALSE, plot = c("var boxplot", "var by n", "log var boxplot", "log var by n", "corrected var boxplot", "corrected var by n", "standard error", "log standard error"), ...) {
    ## Get the model variance
    model_variance <- MCMCglmm.variance(posteriors, scale = relative)
    colnames(model_variance) <- c(name.subsets(data), "residuals")
    ## Get the data variance
    if(!relative) {
        obs_var <- sd(unlist(data$matrix[[1]][, c(data$call$dimensions)]))^2
    }
    var_name <- ifelse(relative, "relative variance", "variance")
    if(missing(group.cols)) {
        group_cols <- rep("grey", ncol(model_variance))
    }
    sample_sizes <- c(size.subsets(data), "residuals" = nrow(data$matrix[[1]]))
    median_variances <- apply(model_variance, 2, median)

@@

```{r, echo = FALSE}
## Plotting all the variance elements
plot.decompose.var(first_model_results, first_model_run, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = 1:8, relative = FALSE)
```

# More complex model with equal sample size

For this model we can use the charadriiformes example data from `dispRity` where the groups are slightly different sizes (but equal-ish) but based on biological data.

@@

```{r, echo = FALSE}
plot.decompose.var(chara_data, charadriiformes$posteriors, group.cols = c(clade_cols, "residuals" = "darkgrey"))
```

# Model with different sample sizes

This is the big model run just on 2D and only with the super orders levels.
The differences with the true results are due to it:
 * not being the same dataset!
 * not having converged...

```{r, echo = FALSE}
## Load the models from the local files
load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain1_20.rda")
model_1_20 <- shapespace_allbirds_lvl_superorder_order_2D
load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain21_40.rda")
model_21_40 <- shapespace_allbirds_lvl_superorder_order_2D
load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain41_60.rda")
model_41_60 <- shapespace_allbirds_lvl_superorder_order_2D
load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain61_80.rda")
model_61_80 <- shapespace_allbirds_lvl_superorder_order_2D
load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain81_100.rda")
model_81_100 <-shapespace_allbirds_lvl_superorder_order_2D
## Combine all the models
model <- combine.mini.chains(unlist(list(model_1_20, model_21_40, model_41_60, model_61_80, model_81_100), recursive = FALSE))

@@

```{r, echo = FALSE}
## Plotting the variance with different data
plot.decompose.var(birds_data, model, group.cols = c(clade_cols, "residuals" = "darkgrey"))
```

# Model used in this paper (8D + 32 terms)

@@

## Set up the dispRity object
full_model <- MCMCglmm.subsets(data       = shapespace,
                               posteriors = model,
                               group      = MCMCglmm.levels(model)[-37],
                               rename.groups = c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny"))

covar.plot(full_model, col = clade_cols, ellipses = mean, legend = TRUE, lines.lty = c(rep(1, length(levels(shapespace$level1)[-1])), rep(2, length(levels(shapespace$level2)[-1])), 1), points = FALSE)
```

```{r, echo = FALSE}
## Plotting the variance with different data
plot.decompose.var(full_model, model, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = 4)
```

>>>>

```{r}
library(dispRity)
library(mcmcmcglmmm)
library(dads)
```
```{r, echo = FALSE}
#@ name plot.decompose.var
#@ description plotting the variance of a model in different ways
#@ param data a dispRity object
#@ param posteriors a MCMCglmm object
#@ param group.cols the colours for each term
#@ param obs.col the colour for the observed variance
#@ param relative whether to make the variance relative or absolute
#@ param plot which panels to plot. Can be any of the followings (default is all):
# "var boxplot" = variance per group;
# "var by n" = variance per sample size;
# "log var boxplot" = logged variance per group;
# "log var by n" = logged variance per sample size;
# "corrected var boxplot" = logged variance divided by logged sample sizel;
# "corrected var by n" = logged variance per logged sample size divided by logged sample size;
# "standard error" = variance per group divided by sqrt(n);
# "log standard error" = logged standard error;
plot.decompose.var <- function(data, posteriors, group.cols, obs.col = "red", relative = FALSE, plot = c("var boxplot", "var by n", "log var boxplot", "log var by n", "corrected var boxplot", "corrected var by n", "standard error", "log standard error"), ...) {
    ## Get the model variance
    model_variance <- MCMCglmm.variance(posteriors, scale = relative)
    colnames(model_variance) <- c(name.subsets(data), "residuals")
    ## Get the data variance
    if(!relative) {
        obs_var <- sum(apply(data$matrix[[1]][, c(data$call$dimensions)], 2, var))
    }
    var_name <- ifelse(relative, "relative variance", "variance")
    if(missing(group.cols)) {
        group_cols <- rep("grey", ncol(model_variance))
    }
    sample_sizes <- c(size.subsets(data), "residuals" = nrow(data$matrix[[1]]))
    median_variances <- apply(model_variance, 2, median)

@@

# Variance explained in a 1D model

For this first very simple model, I will just simulate two trees of 20 taxa each with the two same independent Brownian Motion (BM) traits and combine them together to get a dataset with one variable (2 independent BMs) and two monophyletic clades (`"s"` and `"t"` because why not).

I will run the simplest model which is the data (the 2 BMs) as a response variables, clades as explanatory ones and then two error terms: the data residuals and the phylogenetic randoms terms:

```
data ~ clade + residuals + random(phylogeny)
```

```{r, echo = FALSE, message = FALSE}
test <- MCMCglmm(fixed  = variable ~ clade,
                 random = ~ animal,
                 pedigree = tree,
                 data = data, 
                 nitt = 10000,
                 burnin = 10,
                 thin = 10)
```

I can then extract the variance for the error terms and plot the observed variance in the data (i.e. the variance from the trait)@

```{r, echo = FALSE}
par(mar = c(5, 6, 4, 2) + 0.1)
var_table <- MCMCglmm.variance(test, scale = FALSE)
colnames(var_table) <- c("phylogeny", "residuals")
boxplot(var_table, horizontal = TRUE, las = 1, xlab = "Variance")
abline(v = var(data$variable), col = "red", lwd = 2)
legend("topright", col = "red", lwd = 2, lty = 1, legend = "Observed variance")
```

One first striking thing (to me) is that the residuals are below the observed variance, which is expected: the residuals cannot explain more than the variance in the datasets but this is not the case for the random terms (phylogeny)!

# Variance explained in a 2D model with covariance

We can then spice things up a bit more by looking at what does that variance extracted from the models mean in multi dimensions.
Using the same data as before but adding a random variable (a normal distribution - not related to the phylogeny) we can then do the same model but in a multidimensional way:

The model is now:

```
data(1D, 2D)~ clade + residuals + random(phylogeny)
```

```{r, echo = FALSE, message = FALSE}
## Setting up the model
first_model <- make.mini.chains(data, dimensions = c(1:2), tree = tree, trait.family = "gaussian", residuals = "global", randoms = c("global"), verbose = TRUE, parameters =  list(nitt = 20000))

```{r, echo = FALSE}
par(mar = c(5, 6, 4, 2) + 0.1)
var_table <- MCMCglmm.variance(first_model_run, scale = FALSE)
colnames(var_table) <- c("phylogeny", "residuals")
boxplot(var_table, horizontal = TRUE, las = 1, xlab = "Variance")
abline(v = var(data$variable)+var(data$var2), col = "red", lwd = 2)
legend("topright", col = "red", lwd = 2, lty = 1, legend = "Observed variance")
```

We now see something more funky with the variance of the residuals being sometimes higher than the variance available in the model.
I suspect this is due to the multidimensionality but not sure.

# Variance explained in a 2D model with covariance and nested terms

We can then use the actual model type we used in our paper with a multidimensional variable and a random term for the phylogeny AND each clade:

```
data(1D, 2D) ~ residuals + random(phylogeny) + random(clade_s) + random(clade_t)
```

```{r, echo = FALSE, message = FALSE}
## Setting up the model
second_model <- make.mini.chains(data, dimensions = c(1:2), tree = tree, trait.family = "gaussian", residuals = "global", randoms = c("global", "clade"), verbose = TRUE, parameters =  list(nitt = 20000))

## running the first model
second_model_run <- run.mini.chains(second_model, replicates = 1)[[1]]
```

And this is what it looks like graphically:

```{r, echo = FALSE}
## Model object
second_model_results <- MCMCglmm.subsets(
      data          = data,
      posteriors    = second_model_run,
      group         = MCMCglmm.levels(second_model_run)[c(1:3)],
      rename.groups = c("clade_s", "clade_t", "phylogeny"))

## Plotting the model
covar.plot(second_model_results, col = clade_cols, ellipses = mean, legend = TRUE)
```

We can then extract the variance of the model:

```{r, echo = FALSE, fig.height = 16, fig.width = 4}
## Plotting all the variance elements
par(mar = c(5, 6, 4, 2) + 0.1)
plot.decompose.var(second_model_results, second_model_run, group.cols = c(clade_cols, "residuals" = "darkgrey"), relative = FALSE, plot = "var boxplot")
```

We can see now that the data consistently spreads over the total amount of variance (although the median values remain lower).
To avoid this spread we can log the variance to see what it does:

```{r, echo = FALSE, fig.height = 16, fig.width = 4}
## Plotting all the variance elements
par(mar = c(5, 6, 4, 2) + 0.1)
plot.decompose.var(second_model_results, second_model_run, group.cols = c(clade_cols, "residuals" = "darkgrey"), relative = FALSE, plot = "log var boxplot")
```

Bit clearer but still enigmatic to me (i.e. how can we have more variance explained than variance available).

# Multidimensional nested model with different sample sizes.

Another thing with our models is that they deal with different sample sizes.
We can explore that using a mild example (a tree with three groups of roughly equal size) and a more contrasting example (with orders of magnitudes in differences in sample sizes).

## Similar sample sizes

For this model we can use the charadriiformes example data from `dispRity` where the groups are slightly different sizes (159, 98 and 102 species in each clade).

@@

```{r, echo = FALSE}
par(mar = c(5, 6, 4, 2) + 0.1)
plot.decompose.var(chara_data, charadriiformes$posteriors, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = "var boxplot")
```

It's interesting to see that the variance for the phylogeny is greater than for two of the groups (expected?) but not for the gulls group!
Also, in these datasets, the residuals start to become very small (probably due to most variance being explained by the random terms).
Here's a logged illustration of these variance differences:

```{r, echo = FALSE}
par(mar = c(5, 6, 4, 2) + 0.1)
plot.decompose.var(chara_data, charadriiformes$posteriors, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = "log var boxplot")
```

We can then also plot these differences in variances relative to sample sizes (ignore the lm fit):

```{r, echo = FALSE, fig.height = 4, fig.width = 8}
plot.decompose.var(chara_data, charadriiformes$posteriors, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = c("var by n", "log var by n"))
```

The red circle is the total variance (probably OK to ignore in these plots).

## Different sample sizes

We can then look at the effect of more wildly different sample sizes.
For this test, I'm going to run the BIG whole shapespace dataset in 2D with all the super orders as nested levels.
The sample sizes for the nest random terms then varies from 24 (Mirandortnithes) species to >6k (Telluraves) so we have three orders for magnitude here.

```{r, echo = FALSE, fig.height = 6, fig.width = 6}
## Load the models from the local files
load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain1_20.rda")
model_1_20 <- shapespace_allbirds_lvl_superorder_order_2D
load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain21_40.rda")
model_21_40 <- shapespace_allbirds_lvl_superorder_order_2D
load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain41_60.rda")
model_41_60 <- shapespace_allbirds_lvl_superorder_order_2D
load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain61_80.rda")
model_61_80 <- shapespace_allbirds_lvl_superorder_order_2D
load(file = "../Data/Cluster/shapespace_allbirds_lvl_superorder_order/chains_2D_testing/params_shapespace_2D_9levels_20k_chain81_100.rda")
model_81_100 <-shapespace_allbirds_lvl_superorder_order_2D
## Combine all the models
model <- combine.mini.chains(unlist(list(model_1_20, model_21_40, model_41_60, model_61_80, model_81_100), recursive = FALSE))

@@

Note that the results (ellipses) look kind of different than the ones from the real model.
This is interesting in highlighting the effect of using 8 rather than 2 dimensions and using the interactions of all nested terms (including the orders) rather than just the super orders.
Also the calculations here are rough and the model hasn't converged.

And then we can plot the variance and the logged variance per terms and by sample sizes:

```{r, echo = FALSE, fig.height = 10, fig.width = 10}
## Plotting the variance with different data
par(mar = c(5, 6, 4, 2) + 0.1)
plot.decompose.var(birds_data, model, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = c("var boxplot", "log var boxplot", "var by n", "log var by n"))
```

Here there seems to be a clear relation between sample size and variance: big clades have on average lower variance than smaller ones (nicely illustrated on a log log scale).

We can thus try to "correct" for these sample size issues by using the standard error (or log standard error):

```{r, echo = FALSE, fig.height = 4, fig.width = 8}
## Plotting the variance with different data
par(mar = c(5, 6, 4, 2) + 0.1)
plot.decompose.var(birds_data, model, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = c("standard error", "log standard error"))
```

This shows maybe easier to interpret variations.

Or, maybe more interpretable, we can just "correct" variance by dividing it by the log sample size (instead of the square root of the sample size in the standard error) assuming a log normal relation between variance and sample size (kind of clearer in the big model at the end).

```{r, echo = FALSE, fig.height = 4, fig.width = 8}
## Plotting the variance with different data
par(mar = c(5, 6, 4, 2) + 0.1)
plot.decompose.var(birds_data, model, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = c("corrected var boxplot", "corrected var by n"))
```

# Variance on BIG models

We can then apply these thingies to the big model (with 8 dimensions and 32 nested random terms.
This is what it looks like:

@@

## Set up the dispRity object
full_model <- MCMCglmm.subsets(data       = shapespace,
                               posteriors = model,
                               group      = MCMCglmm.levels(model)[-37],
                               rename.groups = c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny"))
covar.plot(full_model, col = clade_cols, ellipses = mean, legend = TRUE, lines.lty = c(rep(1, length(levels(shapespace$level1)[-1])), rep(2, length(levels(shapespace$level2)[-1])), 1), points = FALSE)
```

And these are all the different ways to illustrate this variance:

```{r, echo = FALSE, fig.height = 32, fig.width = 8}
## Plotting the variance with different data
par(mar = c(5, 6, 4, 2) + 0.1)
plot.decompose.var(full_model, model, group.cols = c(clade_cols, "residuals" = "darkgrey"), plot = c("var boxplot", "log var boxplot", "var by n", "log var by n", "corrected var boxplot", "corrected var by n", "standard error", "log standard error"))
```

Not sure what we get out of all of this but hopefully it gives us some insights on how to talk about the variance partition in the intro.
